---
title: "Report of OptiRTC"
author: "Ziwen Yu"
output: html_document
---

### Overview
This is the report of the OptiRTC product development test. It uses the 5-miniute-interval soil moisture observation and precipitation data to analyze the green roof's hydrologic performance in managing storm water. 

```{r Environment options and data loading, echo=FALSE,warning=FALSE,message=FALSE}
if (!require("pacman")) install.packages("pacman", repos="http://cran.rstudio.com/")
pacman::p_load(rjson, 
               lubridate, 
               dplyr,
               tidyr,
               magrittr,
               zoo,
               pander,
               ggplot2,
               scales)

#Constants
height=6 #in
in2ft=1/12
Ft2_2_m3=0.02831685
Po=0.25
arae=1260 #sqft
Volumn=arae*height*in2ft*Ft2_2_m3  #m3


#Read in Data
#Jsonfile='https://raw.githubusercontent.com/OptiRTC/Product-Dev-Test/master/GreenRoofStorms.json'

#Jsonfile='C:\\Users\\Ziwen.Yu\\Documents\\Class\\OptiRTC\\GreenRoofStorms.json'
JsonData=Jsonfile %>% 
      readLines() %>%
      paste0() %>%
      fromJSON()

Data=do.call(rbind, JsonData)%>%
      as.data.frame() %>%
      lapply(.,unlist)%>%
      data.frame(.,stringsAsFactors =F) %>%
      mutate(Time_UTC=ymd_hms(timeValue_UTC)) %>%
      arrange(Time_UTC) 
```

### Data Cleaning
The data contains `r length(JsonData)` rows of data with `r length(JsonData[[1]])` columns of information about time, rain amount and soil moisture reading from four probes. The time span read from the data is `r (with_tz(min(Data$Time_UTC),'UTC') %--% with_tz(max(Data$Time_UTC),'UTC'))/years(1)` years. The distinct year of the time data is listed below:
```{r Frequent of Years,echo=FALSE}
YearFreq=Data$Time_UTC %>%
      year() %>%
      table() %>%
      data.frame() 

colnames(YearFreq)=c('Year','Freq')

pander(YearFreq, split.cells = 30, split.table = Inf)
```
The year 0001 is obviously a invalid date record for an observation data. The 8 rows of data with time in year 0001 is removed from the further analysis.

```{r Date cleaning,echo=FALSE}
#remove data with year of 0001
Data_Clean=Data %>%
      mutate(Year=year(Time_UTC)) %>%
      filter(Year==2015) %>%
      select(-Year,-timeValue_UTC) %>%
      mutate(ID=as.integer(row.names(.)))
```
After the cleaning of date part, the time period of the remaining data spans `r as.character(min(Data_Clean$Time_UTC) %--% max(Data_Clean$Time_UTC))`.

### Data Exploration
```{r Plot Data,echo=FALSE}

Plot_theme = theme_bw()+
      theme(plot.title = element_text(size = 12)
            ,axis.title.x = element_text(size=10)
            ,axis.title.y = element_text(size=10)
            ,axis.text.x = element_text(size=8,angle = 45, hjust = 1)
            ,axis.text.y = element_text(size=8)            
      )

# Plot Soil Moisture
ggplot(data=Data_Clean)+
      geom_line(aes(x=Time_UTC,y=NorthCentralSoilMoisture_m3_per_m3,color='NorthCentral'),size=1)+
      geom_line(aes(x=Time_UTC,y=SouthCentralSoilMoisture_m3_per_m3,color='SouthCentral'),size=1)+
      geom_line(aes(x=Time_UTC,y=NorthSoilMoisture_m3_per_m3,color='North'),size=1)+
      geom_line(aes(x=Time_UTC,y=SouthSoilMoisture_m3_per_m3,color='South'),size=1)+
      scale_y_continuous(labels = percent_format())+
      labs(y="Soil Moisture",x="",title="Soil Moisture Observations")+
      Plot_theme+
      theme(legend.position="bottom", 
            legend.direction="horizontal",
            legend.title = element_blank())

# Plot Rain
ggplot(data=Data_Clean)+
      geom_bar(aes(x=Time_UTC,y=Rain_in,fill='Rain'),stat="identity")+
      labs(y="Rain (in)",x="Time",title='Rain Observations')+
      Plot_theme+
      theme(legend.position="none")
      
```


### Rain Event Categorization
In defining a rain event, an inter-event dry period is assumed as **4 hours**. In another word, two rain events separated by a dry period no longer than 4 hours are considered one.
```{r Rain Event Define, echo=FALSE,warning=FALSE,message=FALSE}
#---------------Get precipitation events
#select only precipitation hours
# Add previous and post hours to each row
# Calculate the interval previous and post
# Inter-event period =4 hours 
Precip=Data_Clean %>%
      filter(Rain_in>0) %>%
      select(Time_UTC,Rain_in,ID) %>%
      arrange(Time_UTC) %>%
      mutate(Pre=lag(with_tz(Time_UTC,'UTC'),1),
             Post=lead(with_tz(Time_UTC,'UTC'),1))

# add the boundary dry period longer than 4 hours 
# at the beginning and end of rain hours series
EndRw=dim(Precip)[1]
Precip$Pre[1]=with_tz(Precip$Time_UTC[1]-hours(10),'UTC')
Precip$Post[EndRw]=with_tz(Precip$Time_UTC[EndRw]+hours(10),'UTC')

Precip %<>%
      mutate(PreInt=Pre %--% Time_UTC,
             PostInt=Time_UTC %--% Post) %>%
      mutate(PreInt=PreInt/hours(1),
             PostInt=PostInt/hours(1)) 

#  Get the start and end of a rain period
EventSt=Precip %>%
      filter(PreInt>4) %>%
      select(ID)
EventEd=Precip %>%
      filter(PostInt>4) %>%
      select(ID)
Event=data.frame(Start=EventSt$ID,End=EventEd$ID)
Event %<>%
      mutate(Duration=Data_Clean$Time_UTC[Start] %--% Data_Clean$Time_UTC[End]) %>%
      mutate(Length_min=Duration/minutes(1)+5) %>%
      mutate(EvtID=as.integer(row.names(.))) %>%
      cbind(.,EvtRain_in=mapply(function(x,y) sum(Data_Clean$Rain_in[x:y]),Event$Start,Event$End))

pander(Event %>% 
             select(EvtID,Duration,Length_min,EvtRain_in) %>% 
             mutate(Duration=as.character(Duration)), split.cells = 30, split.table = Inf)


# observation realted to rain events
Event_Ob=apply(Event[,c(1,2,5)],1,function(x) Data_Clean %>% filter(ID>=x[1],ID<=x[2]) %>% mutate(EvtID=x[3],EvtSt=x[1]) %>% select(Time_UTC,EvtID))


ggplot(data=Data_Clean %>% left_join(.,do.call(rbind,Event_Ob)))+
      geom_bar(aes(x=Time_UTC,y=Rain_in,fill=factor(EvtID)),stat="identity")+
      labs(y="Rain (in)",x="",title='Rain Event Category',fill='Event ID')+
      Plot_theme+
      theme(legend.position="bottom", 
            legend.direction="horizontal")
```


### Probe Reaction Lag Determination
The observation of soil moisture bears a vibration for all probes. To determine the reaction to a  rain event, a threshold of moisture increase should be pre-defined for each probe from there observations unrelated to any rain event. In this report, the observations before `r as.character(min(Precip$Time_UTC))` are used to conduct this esitmation. 
These thresholds are listed below:
```{r Reading Vibration,echo=FALSE}

#Add delta moisture for all probes

Data_Clean %<>%
      mutate(NorthCentral_lag=lag(NorthCentralSoilMoisture_m3_per_m3,1),
             SouthCentral_lag=lag(SouthCentralSoilMoisture_m3_per_m3,1),
             North_lag=lag(NorthSoilMoisture_m3_per_m3,1),
             South_lag=lag(SouthSoilMoisture_m3_per_m3,1)) %>%
      mutate(NorthCentral_Delta=NorthCentralSoilMoisture_m3_per_m3-NorthCentral_lag,
             SouthCentral_Delta=SouthCentralSoilMoisture_m3_per_m3-SouthCentral_lag,
             North_Delta=NorthSoilMoisture_m3_per_m3-North_lag,
             South_Delta=SouthSoilMoisture_m3_per_m3-South_lag)


# Get threshold increase of moistures to determine 
# the reaction to rain
Event_unrelated=Data_Clean[ Data_Clean$ID < Event$Start[1], ]
Moist_threshold=data.frame(NorthCentral=max(Event_unrelated$NorthCentral_Delta[!is.na(Event_unrelated$NorthCentral_Delta)])
                           ,SouthCentral=max(Event_unrelated$SouthCentral_Delta[!is.na(Event_unrelated$SouthCentral_Delta)])
                           ,North=max(Event_unrelated$North_Delta[!is.na(Event_unrelated$North_Delta)])
                           ,South=max(Event_unrelated$South_Delta[!is.na(Event_unrelated$South_Delta)]))

pander(Moist_threshold, split.cells = 30, split.table = Inf)
```

The reaction lags of all probes for all rain events are tabulated below:
```{r Event react lag,echo=FALSE}
# observation realted to rain events
Event_Ob_delta_Moi=apply(Event[,c(1,2,5)],1,function(x) Data_Clean %>% filter(ID>=x[1],ID<=x[2]) %>% mutate(EvtID=x[3],EvtSt=x[1]))

#------------------ Start of moisture change
#Reaction lag time min
React_lag=function(x)
{
      #NorthCentral
      NorthCentral_React_lag_min=x %>%
            filter(NorthCentral_Delta>Moist_threshold$NorthCentral) %>%
            mutate(React_lag_min=(ID-EvtSt)*5) %>%
            select(React_lag_min) %>%
            min()
      
      #SouthCentral
      SouthCentral_React_lag_min=x %>%
            filter(SouthCentral_Delta>Moist_threshold$SouthCentral) %>%
            mutate(React_lag_min=(ID-EvtSt)*5) %>%
            select(React_lag_min) %>%
            min()
      
      #North
      North_React_lag_min=x %>%
            filter(North_Delta>Moist_threshold$North) %>%
            mutate(React_lag_min=(ID-EvtSt)*5) %>%
            select(React_lag_min) %>%
            min()
      
      #South
      South_React_lag_min=x %>%
            filter(South_Delta>Moist_threshold$South) %>%
            mutate(React_lag_min=(ID-EvtSt)*5) %>%
            select(React_lag_min) %>%
            min()
      
      data.frame(NorthCentral_React_lag_min,
                 SouthCentral_React_lag_min,
                 North_React_lag_min,
                 South_React_lag_min)
}

EventReact=Event %>% cbind(do.call(rbind,lapply(Event_Ob_delta_Moi,React_lag)))

pander(EventReact %>% select(EvtID,
                             NorthCentral_React_lag_min,
                             SouthCentral_React_lag_min,
                             North_React_lag_min,
                             South_React_lag_min), split.cells = 30, split.table = Inf)
```

### Stored Water Estimation
Using the benchmarked soil moisture increase in the last section, the stored water in 15 miniutes is calculated and plotted.
```{r 15min stored water,echo=FALSE,warning=FALSE,message=FALSE,results='hide'}
RunningSumWater= function(x)
{
      x %>%
            #select valid moisture increase reaction for all probes
            mutate(NorthCentral_trueincrease=NorthCentral_Delta*(NorthCentral_Delta>Moist_threshold$NorthCentral),
                   SouthCentral_trueincrease=SouthCentral_Delta*(SouthCentral_Delta>Moist_threshold$SouthCentral),
                   North_trueincrease=North_Delta*(North_Delta>Moist_threshold$North),
                   South_trueincrease=South_Delta*(South_Delta>Moist_threshold$South)) %>%
            #Average the moisture increase
            mutate(AvgMoist_increase=(NorthCentral_trueincrease+
                                            SouthCentral_trueincrease+
                                            North_trueincrease+
                                            South_trueincrease)/4) %>%
            #Calculate water volumn m3
            mutate(stored_water_m3=AvgMoist_increase*Volumn) %>%
            #Running sum of 15 min
            mutate(stored_water_m3_15min=rollsum(x=stored_water_m3,k=3,align = "right",fill=NA)) %>%
            select(Time_UTC,Rain_in,stored_water_m3_15min,EvtID)
}

Stored_water_15min_ls=lapply(Event_Ob_delta_Moi,RunningSumWater)

PlotStoredwater=function(x)
      {      
      ggplot(data=x)+
            geom_line(aes(x=Time_UTC,y=stored_water_m3_15min,color='Stored Water in 15 miniutes (m3)'))+
            geom_bar(aes(x=Time_UTC,y=Rain_in,fill='Rain (in)'),stat="identity")+
            labs(x='',y='',
                 title=paste('Rain Event ID:',unique(x$EvtID)))+
            scale_x_datetime(labels = date_format("%Y-%m-%d %H:%M"))+
            scale_color_manual(values=c('Stored Water in 15 miniutes (m3)'="blue4"))+
            theme_bw()+
            theme(plot.title = element_text(size = 12)
            ,axis.title.x = element_text(size=10)
            ,axis.title.y = element_text(size=10)
            ,axis.text.x = element_text(size=8)
            ,axis.text.y = element_text(size=8)  
            ,legend.position="bottom", 
                  legend.direction="horizontal",
                  legend.title = element_blank()
      )
      }

lapply(Stored_water_15min_ls,PlotStoredwater)

```
